TEMPERATURES AND METALLICITIES OF M DWARFS IN THE APOGEE SURVEY

### 2019-08-01 paper/draft.tex
unitallicize

\title{TEMPERATURES AND METALLICITIES OF M DWARFS IN THE \apogee\ SURVEY}

@@

M dwarfs have enourmous potential for understanding structure and
formation on both Galactic and exoplanetary scales
through their properties and compositions.
However, current atmospheric models
have limited ability to reproduce spectral features in stars at the 
coolest temperatures ($T_{\rm eff} < 4200\,$K) and to fully
exploit the information content of current and upcoming large-scale spectroscopic surveys.
Here we present a catalog of spectroscopic temperatures, metallicities, and
spectral types for 5,875 M dwarfs in the \apogee+\drtwo\ surveys using
\thecannon: a flexible, data-driven spectral-modeling and
parameter-inference framework demonstrated to estimate
stellar-parameter labels ($\teff$, $\logg$, $\feh$, and detailed
abundances) to high precision.
Using a training sample of 87 M dwarfs with optically derived labels spanning 
$2860 < \teff < 4130$\,K calibrated with bolometric temperatures, 
$-0.5 < \feh < 0.5$\,dex calibrated with FGK binary metallicities, 
we train a two-parameter model with predictive accuracy (in cross-validation) 
to 77\,K and 0.09\,dex respectively.
We also train a one-dimensional spectral classification model using 51 M dwarfs
with SDSS optical spectral types ranging from M0 to M9, to predictive accuracy of
0.9 types. 
We find \cannon\ temperatures to be in agreement to within 60\,K
compared to a subsample of 1,702 sources with color-derived temperatures, and 
\cannon\ metallicites to be in agreement to within 0.08\,dex metallicity
compared to a subsample of 15 FGK+M or M+M binaries. 
Finally, our comparison between \cannon\ and \apogee\ pipeline (ASPCAP) labels finds that 
ASPCAP is systematically biased towards reporting higher temperatures and lower metallicities. 
\end{abstract}

@@

Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey \citep[SDSS;][]{Eisenstein:2011,Blanton:2017}, and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope \citep[LAMOST;][]{Zhao:2012} have dramatically increased the sample of known M dwarfs \citep{West:2011,Guo:2015} with catalogs of over 70,000 sources, enabling studies of M dwarf abundances on a Galactic scale.
The Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the SDSS III/IV mission has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}. \apogee\ pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have been determined to precisions 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.

@@

\color{gcolor}{A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}) to developed precise empirical relations (as good as $\sim$0.07\,dex). However these metallicity calibrations are inaccessible for analysis with \apogee, as the wavelength coverage does not include important calibration indices in the $K$-band nor information about the overall spectral shape often used to determine $\teff$ and spectral type. Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by \apogee's resolution.}\color{black}

In this work we build a data-driven model for M dwarf \apogee\ spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model, which employs no line lists or radiative transfer models. 
\thecannon\ is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:methods}). The model in this sense is used to \emph{transfer} labels from spectra for which we know parameters to those which we do not. This data-driven approach effectively circumvents the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (and very accurately measured) \emph{reference labels} possibly measured from other data.

@@

Here we take M-dwarf labels from samples of well-characterized stars that are present in the
\sdssiv\ \apogee\ sample, and use those labels to train a model and label all of the M dwarfs observed by
\sdssiv\ \apogee. One set of labels are physical parameters (effective temperatures and metallicities), the other set of labels are spectral types.

This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the \apogee\ and \gaia\ surveys, as well as previous studies of M dwarfs in \apogee.
Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes the our sample selection and derivation of training parameters.
In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$5,800 sources, and examine the validity of our parameters against color-temperature relations and metallicities of binary pairs.
Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results.

@@

The \apogee\ survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}). For low temperatures ($2800<\teff<3500$\,K), the pipeline uses MARCS plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) ATLAS9 \citep{Castelli:2004} plane-parallel models are used.

\apogee\ is primarily designed to target bright stellar populations, particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013}, with the objective of studying galactic composition and evolution. However numerous cool, main sequence sources have also been observed either as targets proposed by the \apogee\ M dwarf ancilliary survey ($\sim$1,200 sources; \citealt{Desphande:2013}), or serendipitously.

A number of studies out of the M dwarf ancilliary survey have already been conducted to measure reliable fundamental atmospheric parameters and kinematic measurements using spectral synthesis of atmospheric model grids. 
These studies include \citealt{Desphande:2013} and \citealt{Gilhool:2018} which have studied the radial and rotational kinematics for 700+ sources;
\citealt{Souto:2017} and \citealt{Souto:2018} which have modeled three exoplanet-hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining $\teff$/$\logg$/metallicity + 13 elemental abundances;
\citealt{Rajpurohit:2018} which tested BT-Settl \citep{Allard:2012} and MARCS \citep{Gustafsson:2008} model grids on 45 M dwarfs to estimate $\teff$/$\logg$/metallicity;
and \citealt{Skinner:2018} which identified and measured mass ratios and radial velocities for 44 M dwarf spectroscopic binaries.
This work complements existing studies by producing model-independent catalog of spectroscopic temperatures and metallicites to test against model predictions for the entire \apogee\ M dwarf sample, which we quantify to contain at least 10,000 sources to date (DR14).

@@

The spectral type training sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from \sdss. For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.

@@

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/cannon_chi_dist.png}
	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources \citep{ElBadry:2018,Skinner:2018}; \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the \apogee --\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$. We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as ``safe''.}
	\label{fig:chi_dist}
\end{figure*}

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/cmd_selection.png}
	\caption{\gaia\ and \zmass\ color-magnitude cuts for the12,037 sources with $\chi^2<100,000$. Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia. The coordinates for the selected quadrangles 
	$\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ corresponding to ($BP-RP$, $M_G$) for the \gaia\ color-magnitudes shown left, and 
	$\left\{ (.7, 4.3), (1.1, 4.3), (.7, 9), (1.1, 9) \right\}$ corresponding to ($J-K$, $M_K$) for the \zmass\ color-magnitudes shown right. }
	\label{fig:cmd_selection}
\end{figure*}

@@

Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources. From there we applied additional selection criteria, described below, to identify a sample of single, main-sequence M stars, with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries:

@@

\item \textbf{Color-magnitude cuts:} Using \gaia\ and \zmass\ photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiples), and subdwarfs below the main sequence.

@@

\begin{deluxetable*}{cccccccc}
\tablecaption{Metallicities of \apogee\ M dwarfs in wide binaries. Reported in this table are the ASPCAP [M/H] and \cannon\ [Fe/H] of the M dwarfs, as well as the metallicity values of their binary companions. The source/survey of the metallicity and spectral classes of the companions are also given. \label{table:binaries}}
\tabletypesize{\scriptsize} 
\tablehead{\colhead{Mdwarf Gaia ID}      & \colhead{Companion Gaia ID}   & \colhead{Mdwarf [M/H]} & \colhead{Mdwarf [Fe/H]} & \colhead{Companion [Fe/H]} & \colhead{Companion source} & \colhead{Separation (AU)}}
	\startdata
	1476768433933264128 & 1476762627137479552 & -0.18      & 0.01        & 0.10           & APOGEE Cannon M   & 16091      \\
	1476646285062821888 & 1476646250703083264 & -0.02      & 0.18        & 0.27           & APOGEE Cannon M   & 4155       \\
	4008743154907642496 & 4008743086188165248 & -0.20      & 0.02        & 0.09           & APOGEE Cannon M   & 11855      \\
	3053911383155592832 & 3053907221325174144 & -0.25      & -0.15       & -0.15          & APOGEE Cannon M   & 12884      \\
	1475867727751525632 & 1475867727751525504 & -0.25      & -0.15       & -0.13          & APOGEE Cannon M   & 5091       \\
	933708229245319552  & 933708164821511424  & -0.23      & 0.11        & 0.01           & APOGEE Cannon M   & 2245       \\
	3888520938618496768 & 3888520938618115456 & -0.16      & -0.15       & -0.14          & APOGEE Cannon M   & 1128       \\
	3888309828090932608 & 3888309832385988992 & -0.24      & -0.01       & 0.04           & APOGEE Cannon M   & 2499       \\
	374137560388179328  & 374137560388178304  & -0.34      & -0.24       & -0.34          & LAMOST FGK        & 5888       \\
	4031346880591589248 & 4031346914951327104 & -0.33      & -0.25       & -0.18          & LAMOST FGK        & 6637       \\
	3443427533502334592 & 3443427533502335744 & 0.06       & 0.20        & 0.32           & LAMOST FGK        & 3151       \\
	1605991802162861696 & 1605991767803123200 & -0.16      & -0.03       & -0.06          & LAMOST FGK        & 1043       \\
	4418850890305439872 & 4418851263967008000 & -0.18      & 0.11        & 0.27           & APOGEE ASPCAP FGK & 4920       \\
	1381455722290898176 & 1381408782592655616 & -0.13      & -0.02       & -0.08          & APOGEE ASPCAP FGK & 2401       \\
	2687603208839189376 & 2687603311918406528 & -0.12      & 0.00        & -0.03          & APOGEE ASPCAP FGK & 11201    
	\enddata
\end{deluxetable*}

@@

As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combinations of \zmass\ and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures). To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 1,702 sources with both $BV$ photometries measured by APASS, and \zmass\ $JHK$ photometries from \apogee. Figure \ref{fig:teff_comparisons} shows \cannon\ vs. photometric temperatures on the right, and ASPCAP vs. photometric temperatures on the left for each of the 1,702 sources, colored by their respective spectroscopic metallicities.

@@

As a check of our test sample metallicity reliability, we cross-matched our M dwarf final sample with the \citealt{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo. 
In total we found 216 of the \apogee\ M dwarfs to have binary pairs (46 FGK+M, 155 M+M, and 15 WD+M). Out of the 155 M+M pairs, 8 pairs contained both pairs in \apogee. Cross-matching the list of FGK+M dwarf companions with several catalogs/surveys with measured stellar metallicities, we found an additional 7 sources with FGK metallicities from LAMOST \citep{Zhao:2012} and APOGEE (ASPCAP). The metallicity measurements for the 15 M dwarf binaries and their companions are given in Table \ref{table:binaries}, and the overall scatter is $0.08\,$dex--an improvement over the scatter of ASPCAP metallicities which is $0.15\,$dex for these 15 sources.

@@

We trained a data-driven model (\thecannon; \citealt{Ness:2015}) to deliver
high-quality atmospheric parameters ($\teff$ and $\feh$) for M-type dwarf stars
from high-resolution infrared spectra from \apogee.
This work was motivated by the problem that M dwarfs stars are difficult to
model physically; the data are better than the models in important senses.
And indeed we find that our data-driven model is both accurate in the
data domain (as a spectral synthesis model) and precise in the latent domain
(as a tool for deriving physical parameters).
This accuracy and precision is consistent with previous work with
\thecannon\ (\citealt{Ness:2015, Casey:2016, Ho:2017a, Ness:2018}), but
extended to a new regime in spectral type ($\teff$) and surface
gravity.
The primary result of this work is that
we have compiled a catalog of 5,875 M dwarfs with \cannon\ temperatures,
metallicities, spectral types, and six-dimensional kinematics.
These data are provided in Table \ref{table:test_results}.

@@

The model we have developed does have limitations however.
For example, it delivers chi-squared values---goodness-of-fit measures---that
are large; the model is not technically an accurate description of the spectra,
especially when the spectra are observed at signal-to-noise levels above 100.
The model does not include some known physical and instrumental effects.
It does not include terms for line broadening from rotation or
convection (for example, \citealt{Behmard2019}), nor
does it account for binarity or superposition of multiple
stellar spectra (as in, say, \citealt{ElBadry:2018}).
The model also does not include any adjustments for instrumental variations, like for
example the small but significant variations of \apogee\ resolution with
spectrograph fiber number (as included in \citealt{Ness2018}).

The \apogee\ instrument was designed to be sensitive to more than
a dozen individual element abundances in stellar spectra.
So the M-dwarf spectra analyzed here contain individual element abundance
information that we have ignored.
Exploitation of that information requires a better training set of M dwarfs than
we have at present, but is an important goal for the future with these data.

@@

Given that physics-based spectral models of M dwarfs have issues, one of the possible
future values of the data-driven model shown here is that it is highly interpretable:
It contains within it first and second derivatives of the spectral expectation with
respect to the atmospheric parameters.
We show some of these derivatives in Figures \ref{fig:demo_teff}, \ref{fig:demo_feh}, and \ref{fig:west_derivative} and deliver relevant data in
Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}.
These Tables summarize spectral features in the \apogee\ bandpass
that are found to be strong temperature and metallicity indicators.
In the long run, this is the primary value of data-driven models for astronomy:
To provide physical insights that drive physical understandings.
It is our hope that \thecannon, and models like it, will lead to
new and improved physical models which will, in turn, put \thecannon\ out of business. \\

@@

\facilities{
	\sdssiv(\apogee),
	\gaia}

>>>>

\title{TEMPERATURES AND METALLICITIES OF M DWARFS IN THE APOGEE SURVEY}

@@

M dwarfs have enourmous potential for understanding structure and
formation on both Galactic and exoplanetary scales
through their properties and compositions.
However, current atmospheric models
have limited ability to reproduce spectral features in stars at the 
coolest temperatures ($T_{\rm eff} < 4200\,$K) and to fully
exploit the information content of current and upcoming large-scale spectroscopic surveys.
Here we present a catalog of spectroscopic temperatures, metallicities, and
spectral types for 5,875 M dwarfs in the APOGEE+\drtwo\ surveys using
\thecannon: a flexible, data-driven spectral-modeling and
parameter-inference framework demonstrated to estimate
stellar-parameter labels ($\teff$, $\logg$, $\feh$, and detailed
abundances) to high precision.
Using a training sample of 87 M dwarfs with optically derived labels spanning 
$2860 < \teff < 4130$\,K calibrated with bolometric temperatures, 
$-0.5 < \feh < 0.5$\,dex calibrated with FGK binary metallicities, 
we train a two-parameter model with predictive accuracy (in cross-validation) 
to 77\,K and 0.09\,dex respectively.
We also train a one-dimensional spectral classification model using 51 M dwarfs
with SDSS optical spectral types ranging from M0 to M9, to predictive accuracy of
0.9 types. 
We find \cannon\ temperatures to be in agreement to within 60\,K
compared to a subsample of 1,702 sources with color-derived temperatures, and 
\cannon\ metallicites to be in agreement to within 0.08\,dex metallicity
compared to a subsample of 15 FGK+M or M+M binaries. 
Finally, our comparison between \cannon\ and APOGEE pipeline (ASPCAP) labels finds that 
ASPCAP is systematically biased towards reporting higher temperatures and lower metallicities. 
\end{abstract}

@@

Advances in instrumentation with the launch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey \citep[SDSS;][]{Eisenstein:2011,Blanton:2017}, and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope \citep[LAMOST;][]{Zhao:2012} have dramatically increased the sample of known M dwarfs \citep{West:2011,Guo:2015} with catalogs of over 70,000 sources, enabling studies of M dwarf abundances on a Galactic scale.
The Apache Point Observatory Galactic Evolution Experiment (APOGEE; \citealt{Majewski:2015}) survey as part of the SDSS III/IV mission has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}. APOGEE pipeline measurements of $\teff$ and $\feh$ \citep{Perez:2016} for M dwarfs have been determined to precisions 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.

@@

\color{gcolor}{A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star and strong, isolated lines in the spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}) to developed precise empirical relations (as good as $\sim$0.07\,dex). However these metallicity calibrations are inaccessible for analysis with APOGEE, as the wavelength coverage does not include important calibration indices in the $K$-band nor information about the overall spectral shape often used to determine $\teff$ and spectral type. Further, earlier calibrations are generally based on moderate-resolution data \citep[with some exceptions,][]{Neves:2014, Lindgren:2016} that fail to utilize the greater spectral information provided by APOGEE's resolution.}\color{black}

In this work we build a data-driven model for M dwarf APOGEE spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a,Behmard2019}: a fully empirical model, which employs no line lists or radiative transfer models. 
\thecannon\ is a \emph{generative model} which parameterizes the flux at each pixel of a spectrum in terms of a set of stellar labels (a flexible number of parameters chosen by the user; described in more detail in Section \ref{sec:methods}). The model in this sense is used to \emph{transfer} labels from spectra for which we know parameters to those which we do not. This data-driven approach effectively circumvents the challenges of physically modeling the atmosphere of a star (and common issues associated such as incomplete line lists or opacities), provided that we have a subset of spectra in the dataset with known (and very accurately measured) \emph{reference labels} possibly measured from other data.

@@

Here we take M-dwarf labels from samples of well-characterized stars that are present in the
SDSS-IV APOGEE sample, and use those labels to train a model and label all of the M dwarfs observed by
SDSS-IV APOGEE. One set of labels are physical parameters (effective temperatures and metallicities), the other set of labels are spectral types.

This paper is organized as follows: in Section \ref{sec:data} we describe the technical specifications of the data from the APOGEE and \gaia\ surveys, as well as previous studies of M dwarfs in APOGEE.
Section \ref{sec:methods} describes our model implementation using \thecannon\ framework, and Section \ref{sec:sample_selection} describes the our sample selection and derivation of training parameters.
In Section \ref{sec:results} we present our experimental results, evaluate the predictive accuracy of our models, apply our model to a selected test sample of $>$5,800 sources, and examine the validity of our parameters against color-temperature relations and metallicities of binary pairs.
Finally, in Section \ref{sec:discussion} we discuss model performance, future improvements, and implications of our results.

@@

The APOGEE survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the APOGEE Stellar Parameter and Chemical Abundances Pipeline (ASPCAP; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the FERRE code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}). For low temperatures ($2800<\teff<3500$\,K), the pipeline uses MARCS plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) ATLAS9 \citep{Castelli:2004} plane-parallel models are used.

APOGEE is primarily designed to target bright stellar populations, particularly red giants, with de-reddened photometry and color cutoffs of $7 \leq H \leq 13.8$ and $[J-K]_0 \geq 0.5$ \citep{Zasowski:2013}, with the objective of studying galactic composition and evolution. However numerous cool, main sequence sources have also been observed either as targets proposed by the APOGEE M dwarf ancilliary survey ($\sim$1,200 sources; \citealt{Desphande:2013}), or serendipitously.

A number of studies out of the M dwarf ancilliary survey have already been conducted to measure reliable fundamental atmospheric parameters and kinematic measurements using spectral synthesis of atmospheric model grids. 
These studies include \citealt{Desphande:2013} and \citealt{Gilhool:2018} which have studied the radial and rotational kinematics for 700+ sources;
\citealt{Souto:2017} and \citealt{Souto:2018} which have modeled three exoplanet-hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining $\teff$/$\logg$/metallicity + 13 elemental abundances;
\citealt{Rajpurohit:2018} which tested BT-Settl \citep{Allard:2012} and MARCS \citep{Gustafsson:2008} model grids on 45 M dwarfs to estimate $\teff$/$\logg$/metallicity;
and \citealt{Skinner:2018} which identified and measured mass ratios and radial velocities for 44 M dwarf spectroscopic binaries.
This work complements existing studies by producing model-independent catalog of spectroscopic temperatures and metallicites to test against model predictions for the entire APOGEE M dwarf sample, which we quantify to contain at least 10,000 sources to date (DR14).

@@

The spectral type training sample consists of 51 sources, spanning M0-M9 cross-matched from the \citealt{West:2011} (hereafter W11) catalog of 78,841 M dwarfs from SDSS. For each source in the catalog, spectral types were determined both through an automated routine for comparing spectral type templates to data using \textsl{The Hammer} \citep{Covey:2007} and by visual inspection to a reported accuracy of $\pm$1 type.

@@

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/cannon_chi_dist.png}
	\caption{\textit{Left:} the distribution of fits for the training sample and known samples of pre-MS stars (\citealt{Cottaar:2014}) and binary sources \citep{ElBadry:2018,Skinner:2018}; \textit{right:} the distribution of $\chi^2$ fits for all 14,827 sources in the APOGEE--\gaia\ cross match, with color cuts $1<G_{BP}-G_{RP}<6$ and $7.5<M_{G}<20$ and $\varpi>0$. We apply a quality cut of $\chi^2 < 100,000$ to the test sample for those sources we report as ``safe''.}
	\label{fig:chi_dist}
\end{figure*}

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/cmd_selection.png}
	\caption{\gaia\ and 2MASS color-magnitude cuts for the12,037 sources with $\chi^2<100,000$. Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia. The coordinates for the selected quadrangles 
	$\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ corresponding to ($BP-RP$, $M_G$) for the \gaia\ color-magnitudes shown left, and 
	$\left\{ (.7, 4.3), (1.1, 4.3), (.7, 9), (1.1, 9) \right\}$ corresponding to ($J-K$, $M_K$) for the 2MASS color-magnitudes shown right. }
	\label{fig:cmd_selection}
\end{figure*}

@@

Out of the total APOGEE DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \drtwo\ \citep{Brown:2018} and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$ for sources with only positive parallaxes ($\varpi>0$), yielding a sample of 14,827 sources. From there we applied additional selection criteria, described below, to identify a sample of single, main-sequence M stars, with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries:

@@

\item \textbf{Color-magnitude cuts:} Using \gaia\ and 2MASS photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to remove sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiples), and subdwarfs below the main sequence.

@@

\begin{deluxetable*}{cccccccc}
\tablecaption{Metallicities of APOGEE M dwarfs in wide binaries. Reported in this table are the ASPCAP [M/H] and \cannon\ [Fe/H] of the M dwarfs, as well as the metallicity values of their binary companions. The source/survey of the metallicity and spectral classes of the companions are also given. \label{table:binaries}}
\tabletypesize{\scriptsize} 
\tablehead{\colhead{Mdwarf Gaia ID}      & \colhead{Companion Gaia ID}   & \colhead{Mdwarf [M/H]} & \colhead{Mdwarf [Fe/H]} & \colhead{Companion [Fe/H]} & \colhead{Companion source} & \colhead{Separation (AU)}}
	\startdata
	1476768433933264128 & 1476762627137479552 & -0.18      & 0.01        & 0.10           & APOGEE Cannon M   & 16091      \\
	1476646285062821888 & 1476646250703083264 & -0.02      & 0.18        & 0.27           & APOGEE Cannon M   & 4155       \\
	4008743154907642496 & 4008743086188165248 & -0.20      & 0.02        & 0.09           & APOGEE Cannon M   & 11855      \\
	3053911383155592832 & 3053907221325174144 & -0.25      & -0.15       & -0.15          & APOGEE Cannon M   & 12884      \\
	1475867727751525632 & 1475867727751525504 & -0.25      & -0.15       & -0.13          & APOGEE Cannon M   & 5091       \\
	933708229245319552  & 933708164821511424  & -0.23      & 0.11        & 0.01           & APOGEE Cannon M   & 2245       \\
	3888520938618496768 & 3888520938618115456 & -0.16      & -0.15       & -0.14          & APOGEE Cannon M   & 1128       \\
	3888309828090932608 & 3888309832385988992 & -0.24      & -0.01       & 0.04           & APOGEE Cannon M   & 2499       \\
	374137560388179328  & 374137560388178304  & -0.34      & -0.24       & -0.34          & LAMOST FGK        & 5888       \\
	4031346880591589248 & 4031346914951327104 & -0.33      & -0.25       & -0.18          & LAMOST FGK        & 6637       \\
	3443427533502334592 & 3443427533502335744 & 0.06       & 0.20        & 0.32           & LAMOST FGK        & 3151       \\
	1605991802162861696 & 1605991767803123200 & -0.16      & -0.03       & -0.06          & LAMOST FGK        & 1043       \\
	4418850890305439872 & 4418851263967008000 & -0.18      & 0.11        & 0.27           & APOGEE ASPCAP FGK & 4920       \\
	1381455722290898176 & 1381408782592655616 & -0.13      & -0.02       & -0.08          & APOGEE ASPCAP FGK & 2401       \\
	2687603208839189376 & 2687603311918406528 & -0.12      & 0.00        & -0.03          & APOGEE ASPCAP FGK & 11201    
	\enddata
\end{deluxetable*}

@@

As a validation test of our derived temperatures, we perform a comparison between several color-temperature relations from literature, which use combinations of 2MASS and visual band photometries to predict temperatures (similarly to the \citealt{Schmidt:2016} evaluation of ASPCAP temperatures). To obtain visual band magnitudes for a set of sources, we cross-matched the 5,875 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 1,702 sources with both $BV$ photometries measured by APASS, and 2MASS $JHK$ photometries from APOGEE. Figure \ref{fig:teff_comparisons} shows \cannon\ vs. photometric temperatures on the right, and ASPCAP vs. photometric temperatures on the left for each of the 1,702 sources, colored by their respective spectroscopic metallicities.

@@

As a check of our test sample metallicity reliability, we cross-matched our M dwarf final sample with the \citealt{Elbadry:2018b} catalog of $>$50,000 high-confidence, widely separated binaries identified by \drtwo. 
In total we found 216 of the APOGEE M dwarfs to have binary pairs (46 FGK+M, 155 M+M, and 15 WD+M). Out of the 155 M+M pairs, 8 pairs contained both pairs in APOGEE. Cross-matching the list of FGK+M dwarf companions with several catalogs/surveys with measured stellar metallicities, we found an additional 7 sources with FGK metallicities from LAMOST \citep{Zhao:2012} and APOGEE (ASPCAP). The metallicity measurements for the 15 M dwarf binaries and their companions are given in Table \ref{table:binaries}, and the overall scatter is $0.08\,$dex--an improvement over the scatter of ASPCAP metallicities which is $0.15\,$dex for these 15 sources.

@@

We trained a data-driven model (\thecannon; \citealt{Ness:2015}) to deliver
high-quality atmospheric parameters ($\teff$ and $\feh$) for M-type dwarf stars
from high-resolution infrared spectra from APOGEE.
This work was motivated by the problem that M dwarfs stars are difficult to
model physically; the data are better than the models in important senses.
And indeed we find that our data-driven model is both accurate in the
data domain (as a spectral synthesis model) and precise in the latent domain
(as a tool for deriving physical parameters).
This accuracy and precision is consistent with previous work with
\thecannon\ (\citealt{Ness:2015, Casey:2016, Ho:2017a, Ness:2018}), but
extended to a new regime in spectral type ($\teff$) and surface
gravity.
The primary result of this work is that
we have compiled a catalog of 5,875 M dwarfs with \cannon\ temperatures,
metallicities, spectral types, and six-dimensional kinematics.
These data are provided in Table \ref{table:test_results}.

@@

The model we have developed does have limitations however.
For example, it delivers chi-squared values---goodness-of-fit measures---that
are large; the model is not technically an accurate description of the spectra,
especially when the spectra are observed at signal-to-noise levels above 100.
The model does not include some known physical and instrumental effects.
It does not include terms for line broadening from rotation or
convection (for example, \citealt{Behmard2019}), nor
does it account for binarity or superposition of multiple
stellar spectra (as in, say, \citealt{ElBadry:2018}).
The model also does not include any adjustments for instrumental variations, like for
example the small but significant variations of APOGEE resolution with
spectrograph fiber number (as included in \citealt{Ness2018}).

The APOGEE instrument was designed to be sensitive to more than
a dozen individual element abundances in stellar spectra.
So the M-dwarf spectra analyzed here contain individual element abundance
information that we have ignored.
Exploitation of that information requires a better training set of M dwarfs than
we have at present, but is an important goal for the future with these data.

@@

Given that physics-based spectral models of M dwarfs have issues, one of the possible
future values of the data-driven model shown here is that it is highly interpretable:
It contains within it first and second derivatives of the spectral expectation with
respect to the atmospheric parameters.
We show some of these derivatives in Figures \ref{fig:demo_teff}, \ref{fig:demo_feh}, and \ref{fig:west_derivative} and deliver relevant data in
Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}.
These Tables summarize spectral features in the APOGEE bandpass
that are found to be strong temperature and metallicity indicators.
In the long run, this is the primary value of data-driven models for astronomy:
To provide physical insights that drive physical understandings.
It is our hope that \thecannon, and models like it, will lead to
new and improved physical models which will, in turn, put \thecannon\ out of business. \\

@@

\facilities{
	SDSS-IV (APOGEE),
	\gaia}
